---
title: "Programming Assignment 4"
subtitle: "⚔<br/>always remember to remember the template's origin: XARINGAN"
author: "Merlin Balihaxi (/'mɜːːːː.lɪn/)"
institute: "Data Science Class, Rutgers University"
date: "2025-03-25 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["rutgers", "rutgers-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# loading data_raw.csv
```{r}
data_raw <- read.csv("../data_raw/fricatives/fricatives.csv")
head(data_raw)
```
#### codebook (copied from README.md)
- obs: (character), observation number
- s_cog: (numeric), center of gravity measurement in Hz for [s].
- s_skewness: (numeric), skewness measurement for [s].
- sh_cog: (numeric), center of gravity measurement in Hz for [ʃ].
- sh_skewness: (numeric), skewness measurement for [ʃ].

---
# tidying data_raw
The data has two categories, **centre of gravity** and **skewness** of the **phonemes** [s] and [ʃ].

So we need to tidy the data to thre main variables. *phoneme* (s and ʃ), *cog* (numeric) and *skewness* (numeric).

I do not know if *pivot_long()* function can be used twice or not, so I first make it from wide to "all long", then split from it two new values **cog** and **skewness** by *pivot_wider()* function
```{r}
#| warning: FALSE
#| message: FALSE
library(tidyverse)
library(dplyr)
library("ds4ling")
data_tidy <- data_raw |> 
  pivot_longer(
    cols = c('s_cog', 'sh_cog', 's_skewness', 'sh_skewness'),
    names_to = c('phoneme', 'measure'),
    names_sep = "_"
  ) |>
  pivot_wider(
    names_from = measure,
    values_from = value
)
```

---
# writing data_tidy.csv
LONG LIVE **GPT**!

I used write.csv long time ago, yet I forgot the format.

[REVISIT YOUR CODE MORE OFTEN!]
```{r}
#| warning: FALSE
#| message: FALSE
write.csv(data_tidy, "../data_tidy/data_tidy.csv", row.names = FALSE)
head(data_tidy)
```
---

# descriptive statistics
```{r}
des_sta <- function(x) {
  cat("Mean:", mean(x), "\n")
  cat("SD:", sd(x), "\n")
  cat("Min:", min(x), "\n")
  cat("Max:", max(x), "\n")
  cat("Median:", median(x), "\n")
} # currently I do not know how to add *group_by()* in a function.
```
```{r}
des_sta (data_tidy$cog) # descriptive statistics for centre of gravity (Hz)
des_sta (data_tidy$skewness)  # descriptive statistics for skewness
```
---

# boxplot for *centre of gravity* ~ *phoneme*
```{r, fig.width = 6, fig.height = 4}
#| label: boxplot for phoneme and centre of gravity
data_tidy |>
  ggplot() +
  aes(x = phoneme, y = cog) +
  geom_boxplot() +
  labs( x = 'Phoneme', y = 'Centre of Gravity (Hz)') +
  ggtitle("Centre of Gravity by Phoneme")
```
---

# *stat_summary()* for *skewness* ~ *phoneme*
```{r, fig.width = 6, fig.height = 4}
#| label: boxplot for phoneme and skewness
data_tidy |>
  ggplot() +
  aes(x = phoneme, y = skewness) +
  stat_summary(
    fun.data = mean_sdl,
    geom = 'pointrange'
  ) +
  labs( x = 'Phoneme', y = 'skewness') +
  ggtitle('skewness by Phoneme')
```
---

# linear model of *cog* ~ *skewness* for [s]
Firstly, the required data needs to be filtered.

```{r}
s_data <- data_tidy |>
  filter(phoneme == 's') |>
  select(phoneme, skewness, cog)
head(s_data)
```
We might need **ln(centre of gravity)** to decrease the effect of the huge variance of *cog*. The next two pages are on this issue.
---

# linear model of *cog* ~ *skewness* for [s]
```{r}
mod1 <- lm(data = s_data, cog ~ skewness)
summary(mod1)
```
The intercept is not statistically significant (p = 0.161), meaning the mean of *cog* is not different from 0 when *skewness* = 0 ($H_0$ not rejected).
---

# diagnostics I
```{r fig.width = 8, fig.height = 4}
ds4ling::diagnosis(mod1)
```
- The residuals spread widely (plot 1)
- The histogram is not that bell-shaped (plot 2)
- The tails are deviant from the baseline (plot 3)
- I think I am making up stuff, but I insist using *ln(cog)* instead of *cog*

---
# linear model of *ln(cog)* ~ *skewness* for [s]
#### A model after data transformation from *cog* to *ln(cog)*
```{r}
mod2 <- lm(data = s_data, log(cog) ~ skewness)
summary(mod2)
```
---

# scatterplot for *ln(cog)* ~ *skewness* for [s]
```{r fig.width = 6, fig.height = 4}
s_data |>
  ggplot() +
  aes(x = skewness, y = log(cog)) +
  geom_point() +
  geom_smooth(method = 'lm', formula = y ~ x)
```
#### The scatterplot shows that, for **phoneme [s]**, *ln(cog)* increases when *skewness* increases (positively related).

---

# diagnostics II
```{r fig.width = 8, fig.height = 4}
ds4ling::diagnosis(mod2)
```
- Now the model fits better (**really?**)
- The residuals spread more narrowly (plot 1)
- The histogram is nearly bell-shaped (plot 2, **really?**)
- The tails not that deviant from the baseline (plot 3, **not ture!**)
---

# results of model 2 (*mod2*)
A linear model (*mod2*) is fit to examine the relation between *centre of gravity* and *skewness* in terms of **phoneme [s]**. 

The model is **statistically significant** ($F = 125.1$, $p < 0.001$) and the current predictors have explained 72.27% of the variance in *ln(cog)* ($R^2 = 0.7227$, $Adjusted$ $R^2 = 0.7169$).

The model reveals that *skewness* of [s] is **positively related** to the *log value* of its *centre of gravity* and the relation is statistically significant [ $Es = 0.31922$, $SE = 0.02854$, $t = 11.18$, $p < 0.001$]. This means that for phoneme [s], the increase of *skewness* by 1 unit incurs an increase of *ln(cog)* by 0.31922, or an increase of *cog* by 37.6% ($e^{0.31922} = 1.37605$). The *intercept* is statistically significant [ $Es = 7.40746$, $SE = 0.10898$, $t = 67.97$, $p < 0.001$], meaning when *skewness* = 0, the mean of *ln(cog)* is 7.40746 or *cog* is 1648.235 Hz ($e^{7.40746}$).
---

#loading assumptions.csv
```{r}
assumptions <- read.csv("../data_raw/fricatives/assumptions.csv")
head(assumptions)
```
---

# scatterplot for assumptions.csv
```{r fig.width = 5, fig.height = 4}
assumptions |>
  ggplot() +
  aes(x = x, y = y) +
  geom_point()
```
#### It's a **DINoSAUR**! [▶️ Watch on YouTube](https://www.youtube.com/watch?v=8Oz7DG76ibY)
---

# why it is not appropriate to fit a model
```{r fig.width = 9, fig.height = 4}
mod3 <- lm(data = assumptions, y ~ x)
ds4ling::diagnosis(mod3)
```
- visually, it is a dinosaur if we do the scatterplot
- the residuals plot shows the spreading is very random...
- the histogram shows it is not bell-shaped at all
- the QQ-plot shows that the data at two ends highly deviate from the baseline
---

class: center, middle

# Thanks!

slides created via the R package [**xaringan**](https://github.com/yihui/xaringan)
